### 2016-08-31 latex/paper.tex
Removed dynamic collapse from title
First round of edits from Jim

\begin{center}
\fontsize{14}{20}{\bf Buckling and dynamic collapse of the bicycle wheel}
\end{center}

@@

$^\#$ Department of Mechanical and Industrial Engineering\\
Northeastern University\\
360 Huntington Ave., Boston, MA 02115, USA\\
e-mail: j.papadopoulos@neu.edu\\[2.5ex]

@@

\section*{ABSTRACT}
The spokes of a bicycle wheel must be kept under tension to prevent them from buckling under external loads. However, above a certain critical tension, the rim will bend and twist out of its plane into a ``pringle'' or ``taco'' shape. Despite its importance for wheel strength and safety, there is no formula in the literature for the maximum tension a wheel can withstand before buckling.

Here we derive such a formula. The non-dimensionalized buckling tension depends only on two parameters which are identified as the ratio of torsional stiffness to bending stiffness, and the ratio of spoke stiffness to bending stiffness. The number of ``waves'' in the buckled shape is determined by these two stiffness ratios and may in general by greater than 2. Finally, we show that the wheel may be represented by a system of equivalent springs: two series-connected springs for the rim bending and torsion stiffness, connected in parallel with a spring for the spoke system.

\begin{keywords}
bicycle wheel, 
buckling, 
bifurcation,
bicycle safety.
\end{keywords}

\section{INTRODUCTION}
The earliest ``bicycles'' were built in the workshops of carriage makers who adapted the wheel technology of the time. These wheels typically had iron tires, wooden rims and stout wooden spokes which bore the weight of the machine and rider in compression. The spokes needed to be wide enough to prevent lateral buckling, which made them very heavy and limited their diameter. Around 1869, the tension-spoke wheel, an idea proposed in 1808 by Sir George Cayley but never realized, found its first application on the rapidly-developing bicycle \cite{Ackroyd2011a,Clayton1991a}. Tangential-spoking was developed soon thereafter, and the typical bicycle wheel has not changed dramatically in form since then.

The tension-spoke wheel is a prestressed structure and is susceptible to buckling under internal forces. As the spokes are tightened uniformly, the rim deforms radially to accommodate the spoke strain. At a critical tension, the rim buckles out of its initial plane. The post-buckling configuration is generally stable and the original shape of the wheel can often be recovered by reducing the tension. Buckling can also be triggered by external forces in an otherwise stable wheel leading to a release of strain energy. This is an unstable process which generally leads to a permanently ``taco-ed'' wheel.

@@

He did not discuss the problem further, but suggested that wheel failure commonly occurs due to a loss of lateral stability caused by spoke buckling. Pippard and Francis~\cite{Pippard1932d} derived an equation for lateral deflection based on an elastic foundation model, but did not discuss stability and neglected any effects of spoke tension.

The bicycle wheel can be modeled as an elastically supported arch in which the included angle is allowed to go to $2\pi$. Timoshenko and Gere~\cite{Timoshenko1961a} gave a formula for the critical load for a ring with doubly-symmetric cross-section subjected to radial loads. The theory of flexural-torsional buckling of monosymmetric arches (bicycle rims have only one plane of symmetry) was broadly formalized by Trahair and Papangelis~\cite{Trahair1987b} using the virtual work approach to derive the equilibrium and stability equations. Their theory has been extended to treat arches restrained by continuous~\cite{Pi2002b} or discrete~\cite{Bradford2002d} elastic supports and elastic end restraints~\cite{Guo2014b}.

The problem of the prestressed bicycle wheel is unique for a number of reasons. First, buckling is caused by internal rather than external forces. Second, the spokes act both as elastic restraints resisting buckling and as prestressing elements causing buckling. Third, the lateral, radial, and torsional restraining actions of the spokes are coupled---i.e. lateral deflection at a spoke may produce lateral, radial, and torsional reactions. These considerations extend to other structural systems. Large observation wheels such as the London Eye~\cite{Mann2001a} and the Singapore Flyer~\cite{Allsop2009a} resemble large bicycle wheels and achieve lateral stability due to the bracing angle of prestressed cables, and must be designed against flexural-torsional buckling.

@@

To hold the spokes in tension, the rim must be under compression. An accurate approximation\footnote{95\% accurate for 6 spokes, $\alpha=10^{\circ}$, $\beta=10^{\circ}$. Increasing the number of spokes only improves the estimate.} was given by Sharp~\cite{Sharp1977a} by considering a force balance on the top (or bottom) half of the rim.
	\begin{equation}\label{eq:TN}
	N_r = \frac{n_sT}{2\pi}
	\end{equation}
The spokes act like guy-wires, both supporting the rim radially and stabilizing the rim laterally. However, if the spoke tension exceeds a critical value $T_c$, the compression in the rim causes the wheel to buckle laterally, bending and twisting out of its initial plane, as shown in Fig. \ref{fig:schematic} (c).

@@

The rim is modeled as an initially circular beam with a constant cross-section. For simplicity we will assume that the shear center and centroid of the cross-section coincide. Except for warping deformation, plane cross-sections remain planar and transverse shear flexibility is neglected. Furthermore, we assume that the rim radius is much larger than the cross-sectional dimensions.

@@

The deformed configuration of the rim is described by the displacement of the shear center $\bm{u}=u\e + v\ee + w\eee$ and the twist angle $\phi$. The in-plane curvature $\kappa_1$ remains constant during flexural-torsional buckling. The in-plane curvature, out-of-plane curvature and twist rate are
	\begin{equation}\label{eq:k1}
	\kappa_1 = 1/R
	\end{equation}
	\begin{equation}\label{eq:k2}
	\kappa_2 = u'' - \frac{\phi}{R}
	\end{equation}
	\begin{equation}\label{eq:k3}
	\kappa_3 = \phi' + \frac{1}{R} u'
	\end{equation}
The symbol $()'$ indicates a derivative with respect to arc-length $s$. It can easily be verified that Equations \ref{eq:k1}--\ref{eq:k3} simplify to the standard equations for straight beams in the limit $R\rightarrow \infty$. The total strain energy is decomposed into lateral bending, uniform torsion, and warping terms which are related to the curvatures.
% The curvature and twist rate produce a bending moment and twisting moment equal to
	% \begin{equation}\label{eq:M2}
	% M_2 = EI_{11} \kappa_2
	% \end{equation}
	% \begin{equation}\label{eq:M3}
	% M_3 = GJ \kappa_3 + EI_w \kappa_3''
	% \end{equation}
	\begin{equation}\label{eq:Urim}
	U_{rim} = \frac{1}{2} \int_0^{2\pi} \left( EI_{11} \kappa_2^2 + GJ \kappa_3^2 + EI_w (\kappa_3')^2 \right)\, ds
	\end{equation}
The length of a line segment of the rim is assumed to remain constant during buckling. The apparent shortening of an initial line segment $ds$ is given by the difference between $ds$ and $ds'$, the in-plane component of the deformed line segment.
	\begin{equation}\label{eq:ds}
	ds - ds' = ds - \sqrt{ds^2 - \left(\frac{du}{ds}ds\right)^2} \approx \frac{1}{2} (u')^2 \, ds
	\end{equation}
The axial compressive force in the rim moves through this displacement, performing virtual work equal to
	\begin{equation}\label{eq:Vrim}
	V_{rim} = \frac{1}{2} \int_0^{2\pi} N_r (u')^2 \, ds
	\end{equation}

@@

Formalizing the approach of Smith~\cite{Smith1901a} and Pippard~\cite{Pippard1932d}, we will approximate the spoke system as a continuous elastic ``disc'' which exerts a distributed load on the rim. The stiffness per unit length is given by the sum of the spoke stiffnesses in one periodic grouping of $n_p$ spokes, divided by the arc length of the grouping.
	\begin{equation}\label{eq:kbar}
	\bar{\bm{k}} = \frac{1}{2\pi R}\left(\frac{n_s}{n_p}\right) \sum_i^{n_p} \bm{k}_{f, i}
	\end{equation}
	The distributed load $\bar{\bm{f}}$ does an amount of work per unit length equal to
	\begin{equation}\label{eq:w_spokes}
	\bar{w} = \int_0^{\bm{u}_n} \bar{\bm{f}} \cdot d\bm{u}_n = \frac{1}{2} \bm{u}_n \cdot \bar{\bm{k}} \cdot \bm{u}_n
	\end{equation}
The line of action is assumed to pass sufficiently close to the rim shear center that any difference between the displacement of the spoke nipple and the displacement of the shear center can be ignored. Therefore $\bm{u}_n = \bm{u}$. Substituting a lateral buckling displacement $\bm{u}=u\e$ into Eqn. \ref{eq:w_spokes} and integrating over the length of the rim gives the strain energy stored in the spoke system.
	\begin{equation}\label{eq:Us}
	U_{spokes} = \int_0^{2\pi} \bar{w} \, ds = \frac{1}{2} \bar{k}_{uu}u^2
	\end{equation}
The continuum spoke stiffness can be written $\bar{\bm{k}} = \bar{\bm{k}}^{el} + \bar{\bm{k}}^{tens}$, where the second term is proportional to $T$. A very close approximation for the $\bar{k}_{uu}$ component for a symmetrically dished wheel is given by
	\begin{equation}\label{eq:kuu}
	\bar{k}_{uu} = \frac{n_sE_sA_s}{2\pi Rl}\sin^2{\alpha} + \frac{n_s T}{2\pi Rl}
	\end{equation}

\subsection{Total potential energy and buckling criterion}
The wheel becomes unstable when the increase in strain energy is exactly balanced by the virtual work for a perturbing displacement $\delta u, \delta\phi$, i.e. when the total potential energy is zero.
	\begin{equation}\label{eq:TotPot}
	\Pi = U_{rim} + U_{spokes} - V_{rim} = 0
	\end{equation}
We assume a buckled shape of the form
	\begin{equation}\label{eq:modeshape}
	\begin{split}
	u &= \delta u_n \cos{\frac{ns}{R}} \\
	\phi &= \delta\phi_n \cos{\frac{ns}{R}}
	\end{split}
	\end{equation}
Substituting the buckling mode shape Eqn. \ref{eq:modeshape} into Eqns. \ref{eq:Urim}, \ref{eq:Vrim}, and \ref{eq:Us} and substituting into Eqn. \ref{eq:TotPot} yields a buckling criterion for the $n$th mode.
	\begin{multline}\label{eq:ECrit}
	\frac{\pi EI_{11}}{R^3}(n^2 \delta u_n - R\delta\phi_n)^2 + \frac{\pi n^2}{R^2}\left(GJ + \frac{EI_w}{R^2}n^2\right)(\delta u_n-R\delta\phi_n)^2 + \\
	\pi R(\bar{k}_{uu}^{el} + \bar{k}_{uu}^{el})\delta u_n^2 - N_r\frac{\pi n^2}{R}\delta u_n^2=0
	\end{multline}
The torsional stiffness and warping stiffness can be combined into an effective torsional stiffness $\widetilde{GJ} = GJ + n^2EI_w/R^2$. Equation \ref{eq:ECrit} has a quadratic form and can be written in matrix notation:
	\begin{equation}\label{eq:Qform}
	\begin{bmatrix}
	\delta u_n\\\delta\phi_n
	\end{bmatrix} \cdot
	\begin{bmatrix}
	\frac{\partial^2 \Pi}{\partial u_n^2} & \frac{\partial^2 \Pi}{\partial u_n \partial \phi_n}\\
	\frac{\partial^2 \Pi}{\partial u_n\partial\phi_n} & \frac{\partial^2 \Pi}{\partial \phi_n^2}
	\end{bmatrix} \cdot
	\begin{bmatrix}
	\delta u_n\\\delta\phi_n
	\end{bmatrix}
	=0
	\end{equation}
Non-trivial solutions of Eqn. \ref{eq:Qform} exist when the determinant of the matrix vanishes. Expanding the determinant in Eqn. \ref{eq:Qform} and substituting Eqns. \ref{eq:TN} and \ref{eq:kuu} yields a linear equation for the critical buckling tension of the $n$th mode.
	\begin{multline}\label{eq:Tc_eq}
	\left(\frac{2\pi EI_{11}}{R}+\frac{2\pi \widetilde{GJ}n^2}{R} \right) \left( \frac{2\pi \widetilde{GJ}n^2}{R^3} + \frac{2\pi EI_{11}n^4}{R^3} + 2\pi R\bar{k}_{uu}^{el} + \frac{n_sT_{cn}}{l} - \frac{n_sn^2T_{cn}}{R}\right)\\
	- \left( \frac{2\pi EI_{11}n^2}{R^2} + \frac{2\pi \widetilde{GJ}n^2}{R^2} \right)^2 = 0
	\end{multline}
Solving Eqn. \ref{eq:Tc_eq} for $T_{cn}$ yields the tension at which the $n$th mode becomes unstable. The critical buckling tension $T_c$ is the minimum $T_{cn}$ with respect to the mode number $n\geq 2$.

@@

\subsection{Rims with low torsional stiffness}\label{sec:powerlaw_1}
Many bicycle rims are built from thin-walled open channel sections which have very low torsional stiffness compared with their bending stiffness. Replacing the discrete mode number $n$ with a continuous variable $\bar{n}$, we adopt the following ansatz: (a) $\bar{n}^2 \gg 1$ and (b) $\tilde{\mu}\bar{n}^2 \ll 1$. Under these conditions, Eqn. \ref{eq:Tc_nondim} becomes
	\begin{equation}\label{eq:Tcn_small_mu}
	T_{cn} = \frac{2\pi EI_{11}}{n_sR^2} \left( \tilde{\mu}\bar{n}^4 + \frac{\lambda}{\bar{n}^2}\right)
	\end{equation}
Minimizing Eqn. \ref{eq:Tcn_small_mu} with respect to $\bar{n}$ by setting the derivative $dT_{cn}/d\bar{n}=0$, we obtain the scaling law
	\begin{equation}\label{eq:nbar_small_mu}
	\bar{n} = \left(\frac{\lambda}{2\tilde{\mu}} \right)^{1/6}
	\end{equation}
Substituting Eqn. \ref{eq:nbar_small_mu} into Eqn. \ref{eq:Tc_small_mu} yields, after simplification, a power law for $T_c$.
	\begin{equation}\label{eq:Tc_small_mu}
	T_c \approx \frac{11.875}{n_sR^2} \, \left(\bar{k}_{uu}^{el}R^4 \right)^{2/3} (\widetilde{GJ})^{1/3}
	\end{equation}
Equation \ref{eq:Tc_small_mu} suggests that when the torsional stiffness is smaller than the bending stiffness, the buckling behavior is dominated by torsion and does not depend explicitly on the bending stiffness at all. Equation \ref{eq:Tc_small_mu} is shown in Fig. \ref{fig:Tc_nondim} next to Eqn. \ref{eq:Tc_nondim} and finite-element calculations. The power law is strictly less than Eqn. \ref{eq:Tc_nondim} which conveniently gives a conservative design criterion.

\subsection{Spokes much stiffer than the rim}
A second power law regime appears when $\lambda \gg 1$ and $\tilde{\mu} \sim 1$. Under these conditions, Eqn. \ref{eq:Tc_nondim} becomes
	\begin{equation}\label{eq:Tcn_boef}
	T_{cn} = \frac{2\pi EI_{11}}{n_sR^2}\left(\bar{n}^2 + \frac{\lambda}{\bar{n}^2} \right)
	\end{equation}
Following the same procedure in section \ref{sec:powerlaw_1} gives a separate power law, after simplification.
	\begin{equation}\label{eq:Tc_boef}
	T_c \approx \frac{4\pi}{n_s} \left(\bar{k}_{uu}^{el}EI_{11} \right)^{1/2} 
	\end{equation}
Substituting Eqn. \ref{eq:TN} for $T_c$ gives $N_{rc}=2\left(\bar{k}_{uu}^{el}EI_{11}\right)^{1/2}$, which is precisely the buckling load for an infinite, straight beam on an elastic foundation~\cite{Hetenyi1946b}. Figure \ref{fig:Tc_nondim} (c) shows a comparison of Eqn. \ref{eq:Tc_boef} against Eqn. \ref{eq:Tc_nondim} and finite-element calculations.

\section{EQUIVALENT SPRING MODEL}
The lateral stiffness of the wheel can be decomposed into orthogonal modes in the form of Eqn. \ref{eq:modeshape}. Since the Fourier modes are orthogonal, the total strain energy is the sum of the strain energies in each mode.
	\begin{equation}\label{eq:Umode}
	U_{total} = U_0 + U_1 + U_2 + ...
	\end{equation}

@@

% \section{SUB-CRITICAL BUCKLING UNDER LATERAL LOADS}
% ...

\section{FINITE-ELEMENT CALCULATIONS}
We validated our theoretical predictions against non-linear finite-element calculations implemented in ABAQUS\textsuperscript{\textregistered} 6.14\footnote{ABAQUS\textsuperscript{\textregistered} is a registered trademark of Dassault Syst\`emes.}. The spokes and rim were both modeled using 2-node linear beam elements including shear flexibility (element type B31 in the ABAQUS library). The rim was interpolated with sufficient elements such that any given beam segment spanned less than 6$^{\circ}$ of the circle. Each spoke was modeled with 10 beam elements to accurately reproduce local spoke buckling.

A controlled tensioning strain was applied to the spokes by defining a coefficient of thermal expansion of 1.0 for the spokes (but not the rim) and ramping the temperature down to 150\% of the expected critical buckling strain. Mesh imperfections of the form $\cos{ns/R}$ were added to perturb the initial lateral positions of the rim nodes. Modes starting with $n=2$ up to the expected buckling mode were included. The solution was obtained via an implicit integration procedure (ABAQUS Standard) including non-linear geometric effects.

>>>>

\begin{center}
\fontsize{14}{20}{\bf Buckling of the bicycle wheel}
\end{center}

@@

$^\#$ Department of Mechanical and Industrial Engineering\\
Northeastern University\\
360 Huntington Ave., Boston, MA 02115, USA\\
e-mail: j.papadopoulos@northeastern.edu\\[2.5ex]

@@

\section*{ABSTRACT}
Bicycle spokes brace and stiffen the slender rim, but only as long as they do not lose pre-tension under external loads. However, above a certain critical tension, the rim will buckle into a non-planar saddle shape. Despite its importance for wheel strength and safety, there is no formula in the literature for the maximum tension a wheel can withstand before buckling.

Here we derive such a formula. The non-dimensionalized buckling tension depends only on two parameters which are identified as the ratio of torsional stiffness to bending stiffness, and the ratio of lateral spoke bracing stiffness to rim bending stiffness. The precise buckled shape is determined by these two stiffness ratios and the number of positive peaks of the lateral displacement may in general by greater than 2. We show that the wheel may be represented by a system of equivalent springs: two series-connected springs for the rim bending and torsion stiffness, connected in parallel with a spring for the spoke system.

\begin{keywords}
bicycle wheel, 
bicycle safety, 
wheel strength, 
buckling, 
flexural-torsional.
\end{keywords}

\section{INTRODUCTION}
The earliest bicycles were built in the workshops of carriage makers who adapted the wheel technology of the time. These wheels typically had iron tires, wooden rims and stout wooden spokes which bore the weight of the machine and rider in compression~\cite{Herlihy2004a}. The spokes needed to be wide enough to prevent lateral buckling and to keep the rim planar, which made them very heavy and limited their diameter. Around 1869, the tension-spoke wheel, an idea proposed in 1808 by Sir George Cayley but never realized, found its first application on the rapidly-developing bicycle~\cite{Ackroyd2011a,Clayton1991b}. Tangential-spoking was developed soon thereafter, and the typical bicycle wheel has not changed dramatically in form since then.

The tension-spoke wheel is a prestressed structure and is susceptible to buckling under internal forces. As the spokes are tightened uniformly, the rim deforms radially under uniform compression. At a critical tension, the rim buckles out of its initial plane. The post-buckling load-displacement curve is generally stable and the original shape of the wheel can often be recovered by reducing the tension. Buckling can also be triggered by external forces in an otherwise stable wheel leading to a release of strain energy. This process can lead to plastic yielding of the rim which makes wheel repair significantly more difficult.

@@

He did not discuss the problem of buckling under tension further, but proposed that wheel failure commonly occurs due to lateral buckling when external loads have caused spokes to lose tension. Pippard and Francis~\cite{Pippard1932d} derived an equation for lateral deflection based on an elastic foundation model, but did not discuss stability and neglected any effects of spoke tension.

The theory of bicycle wheel buckling is closely related to the theory of flexural-torsional buckling of arches. Timoshenko and Gere~\cite{Timoshenko1961a} gave a formula for the critical load for a ring with doubly-symmetric cross-section subjected to radial loads. The theory of flexural-torsional buckling of monosymmetric arches (bicycle rim cross-sections generally have only one plane of symmetry) was broadly formalized by Trahair and Papangelis~\cite{Trahair1987b} using the virtual work approach to derive the equilibrium and stability equations. Their theory has been extended to treat arches restrained by continuous~\cite{Pi2002b} or discrete~\cite{Bradford2002d} elastic supports and elastic end restraints~\cite{Guo2014b}.

The problem of the prestressed bicycle wheel is unique for a number of reasons. First, buckling is caused by internal rather than external forces. Second, the spokes act both as elastic restraints resisting buckling and as prestressing elements causing buckling. Third, the lateral, radial, and torsional restraining actions of the spokes are coupled---i.e. lateral deflection of the rim may produce lateral, radial, and torsional reactions from the spokes. These considerations extend to other structural systems. Large observation wheels such as the London Eye~\cite{Mann2001a} and the Singapore Flyer~\cite{Allsop2009a} resemble large bicycle wheels with lateral stability due to the bracing angle of prestressed cables, and must be designed against flexural-torsional buckling.

@@

To hold the spokes in tension, the rim must be under compression. An accurate approximation\footnote{95\% accurate for 6 spokes, $\alpha=10^{\circ}$, $\beta=10^{\circ}$, where angles $\alpha$ and $\beta$ are defined in Fig. \ref{fig:schematic}. Increasing the number of spokes only improves the estimate.} was given by Sharp~\cite{Sharp1977a} by considering a force balance on the top (or bottom) half of the rim.
	\begin{equation}\label{eq:TN}
	N_r = \frac{n_sT}{2\pi}
	\end{equation}
The spokes act like guy-wires, both supporting the rim radially and stabilizing the rim laterally. However, if the spoke tension exceeds a critical value $T_c$, the compression in the rim causes the wheel to buckle laterally, bending and twisting out of its initial plane, as shown in Fig. \ref{fig:schematic} (c).

@@

The rim is modeled as an initially circular beam with a constant cross-section. For simplicity in this abbreviated treatment, we will assume that the shear center and centroid of the cross-section coincide. Except for warping deformation, plane cross-sections remain planar and transverse shear flexibility is neglected. Furthermore, we assume that the rim radius is much larger than the cross-sectional dimensions.

@@

The deformed configuration of the rim is described by the displacement of the shear center $\bm{u}=u\e + v\ee + w\eee$ and the twist angle $\phi$. The in-plane curvature $\kappa_1$ remains constant during flexural-torsional buckling. The in-plane curvature, out-of-plane curvature and twist rate are
	\begin{equation}\label{eq:k1}
	\kappa_1 = 1/R
	\end{equation}
	\begin{equation}\label{eq:k2}
	\kappa_2 = u'' - \frac{\phi}{R}
	\end{equation}
	\begin{equation}\label{eq:k3}
	\kappa_3 = \phi' + \frac{1}{R} u'
	\end{equation}
The symbol $()'$ indicates a derivative with respect to arc-length $s$ along the \eee{} direction. It can easily be verified that Equations \ref{eq:k1}--\ref{eq:k3} simplify to the standard equations for straight beams in the limit $R\rightarrow \infty$. The total strain energy is decomposed into lateral bending, uniform torsion, and warping terms which are related to the curvatures.
% The curvature and twist rate produce a bending moment and twisting moment equal to
	% \begin{equation}\label{eq:M2}
	% M_2 = EI_{11} \kappa_2
	% \end{equation}
	% \begin{equation}\label{eq:M3}
	% M_3 = GJ \kappa_3 + EI_w \kappa_3''
	% \end{equation}
	\begin{equation}\label{eq:Urim}
	U_{rim} = \frac{1}{2} \int_0^{2\pi} \left( EI_{11} \kappa_2^2 + GJ \kappa_3^2 + EI_w (\kappa_3')^2 \right)\, ds
	\end{equation}
The warping energy is primarily related to the bending energy in the sidewalls (which act as beam flanges) of the rim\footnote{See http://www.steel-insdag.org/TeachingMaterial/Chapter17.pdf for a simple treatment.}. The warping constant $I_w$ has units of m$^6$, and for single-wall rims the warping stiffness $EI_w$ generally exceeds the torsion stiffness $GJ$.

The length of a line segment of the rim is assumed to remain constant during buckling. The apparent shortening of an initial line segment $ds$ is given by the difference between $ds$ and $ds'$, the in-plane component of the deformed line segment.
	\begin{equation}\label{eq:ds}
	ds - ds' = ds - \sqrt{ds^2 - \left(\frac{du}{ds}ds\right)^2} \approx \frac{1}{2} (u')^2 \, ds
	\end{equation}
The axial compressive force in the rim moves through this displacement, performing virtual work equal to
	\begin{equation}\label{eq:Vrim}
	V_{rim} = \frac{1}{2} \int_0^{2\pi} N_r (u')^2 \, ds
	\end{equation}

@@

Formalizing the approach of Smith~\cite{Smith1901a} and Pippard~\cite{Pippard1932d}, we will approximate the spoke system as a continuous elastic ``disc'' which exerts a distributed load on the rim. The stiffness per unit length is given by the sum of the spoke stiffnesses in one periodic grouping of $n_p$ spokes, divided by the arc length of the grouping\footnote{For a conventional cross-laced wheel, $n_p=4$: drive-side pulling spoke, drive-side trailing spoke, left-side pulling spoke, left-side trailing spoke. For a radial-spoked wheel, $n_p=2$: drive-side spoke, left-side spoke.}.
	\begin{equation}\label{eq:kbar}
	\bar{\bm{k}} = \frac{1}{2\pi R}\left(\frac{n_s}{n_p}\right) \sum_i^{n_p} \bm{k}_{f, i}
	\end{equation}
	The distributed load $\bar{\bm{f}}$ does an amount of work per unit length equal to
	\begin{equation}\label{eq:w_spokes}
	\bar{w} = \int_0^{\bm{u}_n} \bar{\bm{f}} \cdot d\bm{u}_n = \frac{1}{2} \bm{u}_n \cdot \bar{\bm{k}} \cdot \bm{u}_n
	\end{equation}
Each spoke is assumed to terminate at the rim shear center. Therefore $\bm{u}_n = \bm{u}$. Substituting a lateral buckling displacement $\bm{u}=u\e$ into Eqn. \ref{eq:w_spokes} and integrating over the length of the rim gives the strain energy stored in the spoke system.
	\begin{equation}\label{eq:Us}
	U_{spokes} = \int_0^{2\pi} \bar{w} \, ds = \frac{1}{2} \bar{k}_{uu}u^2
	\end{equation}
The continuum spoke stiffness can be written $\bar{\bm{k}} = \bar{\bm{k}}^{el} + \bar{\bm{k}}^{tens}$, where the second term is proportional to $T$. A very close approximation for the $\bar{k}_{uu}$ component for a symmetrically dished wheel is given by
	\begin{equation}\label{eq:kuu}
	\bar{k}_{uu} = \frac{n_sE_sA_s}{2\pi Rl}\sin^2{\alpha} + \frac{n_s T}{2\pi Rl}
	\end{equation}

\subsection{Total potential energy and buckling criterion}
The wheel becomes unstable when the increase in strain energy is exactly balanced by the virtual work for a perturbing displacement $\delta u, \delta\phi$, i.e. when the total potential energy is zero.
	\begin{equation}\label{eq:TotPot}
	\Pi = U_{rim} + U_{spokes} - V_{rim} = 0
	\end{equation}
We assume a buckled shape of the form
	\begin{equation}\label{eq:modeshape}
	\begin{split}
	u &= \delta u_n \cos{\frac{ns}{R}} \\
	\phi &= \delta\phi_n \cos{\frac{ns}{R}}
	\end{split}
	\end{equation}
\,\,\,\, where $n$ is the number of positive peaks of the lateral displacement. Substituting the buckling mode shape Eqn. \ref{eq:modeshape} into Eqns. \ref{eq:Urim}, \ref{eq:Vrim}, and \ref{eq:Us} and substituting into Eqn. \ref{eq:TotPot} yields a buckling criterion for the $n$th mode.
	\begin{multline}\label{eq:ECrit}
	\frac{\pi EI_{11}}{R^3}(n^2 \delta u_n - R\delta\phi_n)^2 + \frac{\pi n^2}{R^2}\left(GJ + \frac{EI_w}{R^2}n^2\right)(\delta u_n-R\delta\phi_n)^2 + \\
	\pi R(\bar{k}_{uu}^{el} + \bar{k}_{uu}^{el})\delta u_n^2 - N_r\frac{\pi n^2}{R}\delta u_n^2=0
	\end{multline}
The torsional stiffness and warping stiffness can be combined into an effective torsional stiffness $\widetilde{GJ} = GJ + n^2EI_w/R^2$. Equation \ref{eq:ECrit} has a quadratic form and can be written in matrix notation:
	\begin{equation}\label{eq:Qform}
	\begin{bmatrix}
	\delta u_n\\\delta\phi_n
	\end{bmatrix} \cdot
	\begin{bmatrix}
	\frac{\partial^2 \Pi}{\partial u_n^2} & \frac{\partial^2 \Pi}{\partial u_n \partial \phi_n}\\
	\frac{\partial^2 \Pi}{\partial u_n\partial\phi_n} & \frac{\partial^2 \Pi}{\partial \phi_n^2}
	\end{bmatrix} \cdot
	\begin{bmatrix}
	\delta u_n\\\delta\phi_n
	\end{bmatrix}
	=0
	\end{equation}
Non-trivial solutions of Eqn. \ref{eq:Qform} exist when the determinant of the matrix vanishes. Expanding the determinant in Eqn. \ref{eq:Qform} and substituting Eqns. \ref{eq:TN} and \ref{eq:kuu} yields a linear equation for the critical buckling tension of the $n$th mode.
	\begin{multline}\label{eq:Tc_eq}
	\left(\frac{2\pi EI_{11}}{R}+\frac{2\pi \widetilde{GJ}n^2}{R} \right) \left( \frac{2\pi \widetilde{GJ}n^2}{R^3} + \frac{2\pi EI_{11}n^4}{R^3} + 2\pi R\bar{k}_{uu}^{el} + \frac{n_sT_{cn}}{l} - \frac{n_sn^2T_{cn}}{R}\right)\\
	- \left( \frac{2\pi EI_{11}n^2}{R^2} + \frac{2\pi \widetilde{GJ}n^2}{R^2} \right)^2 = 0
	\end{multline}
Solving Eqn. \ref{eq:Tc_eq} for $T_{cn}$ yields the tension at which the $n$th mode becomes unstable. The critical buckling tension $T_c$ is the minimum $T_{cn}$ with respect to the mode number $n\geq 2$.

@@

\subsection{Rims with low torsional stiffness}\label{sec:powerlaw_1}
Many bicycle rims are built from thin-walled open channel sections which have very low torsional stiffness $\widetilde{GJ}$ compared with their bending stiffness. Replacing the discrete mode number $n$ with a continuous variable $\bar{n}$, we adopt the following ansatz: (a) $\bar{n}^2 \gg 1$ and (b) $\tilde{\mu}\bar{n}^2 \ll 1$. Under these conditions, Eqn. \ref{eq:Tc_nondim} becomes
	\begin{equation}\label{eq:Tcn_small_mu}
	T_{cn} = \frac{2\pi EI_{11}}{n_sR^2} \left( \tilde{\mu}\bar{n}^4 + \frac{\lambda}{\bar{n}^2}\right)
	\end{equation}
Minimizing Eqn. \ref{eq:Tcn_small_mu} with respect to $\bar{n}$ by setting the derivative $dT_{cn}/d\bar{n}=0$, we obtain the scaling law
	\begin{equation}\label{eq:nbar_small_mu}
	\bar{n} = \left(\frac{\lambda}{2\tilde{\mu}} \right)^{1/6}
	\end{equation}
Substituting Eqn. \ref{eq:nbar_small_mu} into Eqn. \ref{eq:Tcn_small_mu} yields a power law for $T_c$.
	\begin{equation}\label{eq:Tc_small_mu}
	T_c \approx \frac{11.875}{n_sR^2} \, \left(\bar{k}_{uu}^{el}R^4 \right)^{2/3} (\widetilde{GJ})^{1/3}
	\end{equation}
Equation \ref{eq:Tc_small_mu} suggests that when the torsional stiffness is smaller than the bending stiffness, the buckling behavior is dominated by torsion and does not depend explicitly on the bending stiffness at all. Equation \ref{eq:Tc_small_mu} is shown in Fig. \ref{fig:Tc_nondim} next to Eqn. \ref{eq:Tc_nondim} and finite-element calculations. The power law is strictly less than Eqn. \ref{eq:Tc_nondim} which conveniently gives a conservative design criterion.

\subsection{Lateral stiffness of spoke system much greater than rim bending stiffness}
A second power law regime appears when $\lambda \gg 1$ and $\tilde{\mu} \sim 1$. Under these conditions, Eqn. \ref{eq:Tc_nondim} becomes
	\begin{equation}\label{eq:Tcn_boef}
	T_{cn} = \frac{2\pi EI_{11}}{n_sR^2}\left(\bar{n}^2 + \frac{\lambda}{\bar{n}^2} \right)
	\end{equation}
Following the same procedure in section \ref{sec:powerlaw_1} gives a separate power law, after simplification.
	\begin{equation}\label{eq:Tc_boef}
	T_c \approx \frac{4\pi}{n_s} \left(\bar{k}_{uu}^{el}EI_{11} \right)^{1/2} 
	\end{equation}
Substituting Eqn. \ref{eq:TN} for $T_c$ gives $N_{rc}=2\left(\bar{k}_{uu}^{el}EI_{11}\right)^{1/2}$, which is precisely the buckling load for an infinite, straight beam on an elastic foundation~\cite{Hetenyi1946b}. Figure \ref{fig:Tc_nondim} (c) shows a comparison of Eqn. \ref{eq:Tc_boef} against Eqn. \ref{eq:Tc_nondim} and finite-element calculations.

\section{EQUIVALENT SPRING MODEL}
The form of Eqn. \ref{eq:Tc_nondim} suggests that the buckling tension is governed by an equivalent stiffness which combines the bending stiffness and torsion stiffness of the rim and the lateral stiffness of the spoke system. Here we will describe an equivalent system of linear springs (Fig. \ref{fig:Kn}) consisting of two springs connected in series representing the torsion and bending stiffnesses of the rim, in parallel with a single spring representing the lateral stiffness of the spoke system.

The lateral stiffness of the wheel can be decomposed into orthogonal modes in the form of Eqn. \ref{eq:modeshape}. Since the Fourier modes are orthogonal, the total strain energy is the sum of the strain energies in each mode.
	\begin{equation}\label{eq:Umode}
	U_{total} = U_0 + U_1 + U_2 + ...
	\end{equation}

@@

\section{FINITE-ELEMENT CALCULATIONS}
We validated our theoretical predictions against non-linear finite-element calculations implemented in ABAQUS\textsuperscript{\textregistered} 6.14\footnote{ABAQUS\textsuperscript{\textregistered} is a registered trademark of Dassault Syst\`emes.}. The spokes and rim were both modeled using 2-node linear beam elements including shear flexibility (element type B31 in the ABAQUS library). The rim was interpolated with sufficient elements such that any given beam segment spanned less than 6$^{\circ}$ of the circle (e.g. 2 rim elements between each spoke for a 36-spoke wheel). Each spoke was modeled with 10 beam elements to accurately reproduce local spoke buckling.

A controlled tensioning strain was applied to the spokes by defining a coefficient of thermal expansion of 1.0 for the spokes (but not the rim) and ramping the temperature down to 150\% of the expected critical buckling strain. The initial rim shape was given a geometric imperfection by perturbing the lateral coordinate of the rim nodes by $\cos{ns/R}$. Imperfections starting with $n=2$ up to the expected buckling mode were included. The solution was obtained via an implicit integration procedure (ABAQUS Standard) including non-linear geometric effects.
